Comparison of Fast Imaging Architectures for Large, Multi-scale, Hierarchical Aperture Arrays

### 2024-06-12 ms.tex
Update on Overleaf.

The variable and transient radio sky encompasses a variety of intriguing astrophysical phenomena. These events occur across timescales spanning orders of magnitude, frequency ranges, and polarizations \citep[][and references therein]{Pietka+2015,Chandra+2016}. Notable examples include pulsars and their giant radio pulses, magnetars, fast radio bursts (FRBs), rotating radio transients (RRATs), Jovian and solar bursts, flare stars, cataclysmic variables (CVs), exoplanet and stellar outbursts, X-ray binaries, novae, supernovae, gamma-ray bursts (GRBs) and their afterglows, active galactic nuclei (AGNs) and blazars, tidal disruption events, both the prompt and afterglow counterparts to gravitational wave (GW) merger events, as well as many phenomena yet to be understood \citep{Lorimer+2007,Thyagarajan+2011,Keane2013,Thornton+2013,Bochenek+2020}.  The effectiveness of discovering these transients depends on an instrument's field of view and the duration of on-sky observation \citep{Cordes2007}. Ideally, instruments should operate continuously, covering a wide range of timescales from microseconds to days, with a large instantaneous field of view.

@@

Based on the requirements for transients and observational cosmology, radio interferometers are witnessing a shift to a paradigm of packing a large number of collecting elements in a compact area with high density and large field of view. Modern radio arrays tend to be aperture arrays that operate interferometrically and hierarchically on multiple spatial scales. Examples of these multi-scale hierarchical aperture arrays include the Murchison Widefield Array \cite[MWA;][]{Tingay+2013}, the Hydrogen Epoch of Reionization Array \citep[HERA;][]{HERA+2017}, the Low Frequency Array \cite[LOFAR;][]{vanHaarlem+2013}, the swarm of Long Wavelength Arrays \cite[LWA Swarm;][]{Dowell+2018}, the Canadian Hydrogen Intensity Mapping Experiment \citep[CHIME;][]{CHIME+2022}, and the Square Kilometre Array low \cite[SKA-low;][]{Dewdney+2009,SKA1+2019}.

This paradigm poses severe computational and data rate challenges necessitated by the processing of large numbers of data streams in very short cadence intervals. Unfortunately, compromises are often made by limiting one or more discovery parameters such as time resolution, on-sky observing time, field of view, angular resolution, or bandwidth.

The paper is organised as follows. In \S\ref{sec:notation}, the mathematical notation describing the measurements and imaging process is established. \S\ref{sec:multi-scale-arrays} describes the parametrisation of multi-scale aperture arrays studied in this paper. \S\ref{sec:computational-cost} describes the metric of computational cost density in discovery phase space based on which we compare different imaging architectures described in \S\ref{sec:img-archs}. The multi-scale imaging architecture options for intra- and inter-station data processing are described in sections \ref{sec:inter-station-arch} and \ref{sec:inter-station-arch}, respectively. \S\ref{sec:results} presents a summary of the results, and conclusions are presented in \S\ref{sec:conclusion}.

@@

The specific aperture array examples we consider are the \texttt{SKA-low}, core of the SKA-low (\texttt{SKA-low-core}), \texttt{LAMBDA-I}, \texttt{CASPA} \citep{Luo+2024}, and \texttt{FarView}. The last three are hypothetical interferometers. 
% It is assumed that $N_{\textrm{a/s}_m}=N_\textrm{a/s}=256$ (same for each station), $D_\textrm{a}/\lambda=1$, and $D_\textrm{s}/\lambda=17.5$ for each of these arrays, where, the nominal wavelength chosen here is $\lambda=2$~m. The corresponding station filling factor, $f_\textrm{s}=N_\textrm{a/s}\,(D_\textrm{a}/D_\textrm{s})^2$, is 0.836 for all the arrays. ~\texttt{SKA-low-core}, ~\texttt{SKA-low}, ~\texttt{Aus-VLBA-low}, and ~\texttt{IC-VLBA-low} are taken to consist of $N_\textrm{s}=256, 512, 512,$ and 512 stations in an array diameter of $D_\textrm{A}/\lambda=525, 3.25\times 10^4, 1.95\times 10^6$, and $3.9\times 10^6$, respectively. The corresponding array filling factors, defined as $f_\textrm{A}=N_\textrm{s}\,(D_\textrm{s}/D_\textrm{A})^2$, are 0.284, $1.48\times 10^{-4}$, $4.12\times 10^{-8}$, and $1.03\times 10^{-8}$, respectively. 
These parameters for the chosen examples are summarised in Table~\ref{tab:array_params}.

@@

Fast imaging involves processing data on multiple time scales. The channelisation of a voltage time stream of length $\delta t$ with a time resolution of $t_\textrm{s}$ will correspond to spectral channels of width $\delta\nu=1/\delta t$ and a bandwidth of $B=1/t_\textrm{s}$. The number of spectral channels is $N_\nu=B/\delta\nu=\delta t/t_\textrm{s}$. Images are expected to be produced at a cadence of $t_\textrm{acc}$, which is typically larger than $\delta t$ and smaller than the time scale on which calibration needs to be updated. In this paper, the nominal values chosen for $\delta t$ is 25~$\mu$s, and we explore the possibilities for $t_\textrm{acc}$ ranging from 100~$\mu$s to 10~s.

@@

>>>>

The variable and transient radio sky encompasses a variety of intriguing astrophysical phenomena. These events occur across timescales spanning orders of magnitude, frequency ranges, and polarizations \citep[][and references therein]{Pietka+2015,Chandra+2016}. Notable examples include pulsars and their giant radio pulses, magnetars, fast radio bursts (FRBs), rotating radio transients (RRATs), Jovian and solar bursts, flare stars, cataclysmic variables (CVs), exoplanet and stellar outbursts, X-ray binaries, novae, supernovae, gamma-ray bursts (GRBs) and their afterglows, active galactic nuclei (AGNs) and blazars, tidal disruption events, both the prompt and afterglow counterparts to gravitational wave (GW) merger events, as well as many phenomena yet to be understood \citep{Lorimer+2007,Thyagarajan+2011,Keane2013,Thornton+2013,Bochenek+2020}.  The effectiveness of discovering these transients depends on an instrument's field of view and the duration of on-sky observation \citep{Cordes2007}. Ideally, instruments should operate continuously and in real time, covering a wide range of timescales from microseconds to days, with a large instantaneous field of view.

@@

Based on the requirements for transients and observational cosmology, radio interferometers are witnessing a shift to a paradigm of packing a large number of collecting elements in a compact area with high density and large field of view. Modern radio arrays tend to be aperture arrays that operate interferometrically and hierarchically on multiple spatial scales. Examples of these multi-scale hierarchical aperture arrays include the Murchison Widefield Array \cite[MWA;][]{Tingay+2013}, the Hydrogen Epoch of Reionization Array \citep[HERA;][]{HERA+2017}, the Low Frequency Array \cite[LOFAR;][]{vanHaarlem+2013}, the swarm of Long Wavelength Arrays \cite[LWA Swarm;][]{Dowell+2018}, the Canadian Hydrogen Intensity Mapping Experiment \citep[CHIME;][]{CHIME+2022}, the Hydrogen Intensity and Real-time Analysis Experiment \citep[HIRAX;][]{HIRAX+2022}, and the Square Kilometre Array low \cite[SKA-low;][]{Dewdney+2009,SKA1+2019}.

This new paradigm poses severe computational and data rate challenges necessitated by the processing of large numbers of data streams in very short cadence intervals. The choice of data processing architecture is usually sensitive to the array layout, cadence, angular resolution, and field of view coverage. For example, a correlator-based architecture is generally more suited for relatively smaller number of interferometer elements sparsely distributed over a large area and processing on a slower cadence, whereas a direct imaging architecture based on Fast Fourier Transform (FFT) have far greater efficiency for regularly arranged elements \citep{Daishido+1991,Otobe+1994,Tegmark+2009,Tegmark+2010,Foster+2014,Masui+2019}. A generalised version of FFT-based direct imaging like the E-field Parallel Imaging ``Correlator'' \citep[EPIC;][]{Thyagarajan+2017,Thyagarajan+2019,Krishnan+2023} are found to be highly efficient for large-$N$, densely packed arrays with or without regularly placed elements from computational and data rate standpoints. Pre- or post-correlation beamforming is expected to be efficient when the number of elements and the number of locations where the image is sought are few in number. Thus, the choice of an optimal imaging architecture is critical for maximising the probability of discovery. When the chosen architecture is sub-optimal, compromises are often made by limiting one or more discovery parameters such as time resolution, on-sky observing time, bandwidth, field of view, or angular resolution \citep[for example,][]{Price2024}. When considering multi-scale arrays that involve hierarchical data processing at different scales, there may not a single architecture that is optimal at levels. A hybrid processing architecture that is optimal at various levels of data processing hierarchy is required.

In this paper, our primary motivation is to explore different imaging architectures and their combinations appropriate for a multi-dimensional space of array layout parameters and cadence intervals from a computational standpoint. Conversely, the paper also provides a guide to choosing the array layout parameters and cadence intervals where a given architecture or combinations thereof will remain computationally efficient. Several existing and hypothetical multi-scale aperture arrays are used as examples.

The paper is organised as follows. In \S\ref{sec:notation}, the mathematical notation describing the measurements and imaging process is established. \S\ref{sec:multi-scale-arrays} describes the parametrisation of multi-scale aperture arrays studied. \S\ref{sec:computational-cost} describes the metric of computational cost density in discovery phase space based on which we compare different imaging architectures described in \S\ref{sec:img-archs}. The multi-scale imaging architecture options are described in sections \ref{sec:inter-station-arch} and \ref{sec:inter-station-arch}, respectively. \S\ref{sec:results} contains a discussion and summary of the results, and conclusions are presented in \S\ref{sec:conclusion}.

@@

The specific aperture array examples we consider are the \texttt{SKA-low}, core of the SKA-low (\texttt{SKA-low-core}), \texttt{LAMBDA-I}, \texttt{CASPA} \citep{Luo+2024}, and \texttt{FarView}. The last three are hypothetical interferometers that are being planned for the future. 
% It is assumed that $N_{\textrm{a/s}_m}=N_\textrm{a/s}=256$ (same for each station), $D_\textrm{a}/\lambda=1$, and $D_\textrm{s}/\lambda=17.5$ for each of these arrays, where, the nominal wavelength chosen here is $\lambda=2$~m. The corresponding station filling factor, $f_\textrm{s}=N_\textrm{a/s}\,(D_\textrm{a}/D_\textrm{s})^2$, is 0.836 for all the arrays. ~\texttt{SKA-low-core}, ~\texttt{SKA-low}, ~\texttt{Aus-VLBA-low}, and ~\texttt{IC-VLBA-low} are taken to consist of $N_\textrm{s}=256, 512, 512,$ and 512 stations in an array diameter of $D_\textrm{A}/\lambda=525, 3.25\times 10^4, 1.95\times 10^6$, and $3.9\times 10^6$, respectively. The corresponding array filling factors, defined as $f_\textrm{A}=N_\textrm{s}\,(D_\textrm{s}/D_\textrm{A})^2$, are 0.284, $1.48\times 10^{-4}$, $4.12\times 10^{-8}$, and $1.03\times 10^{-8}$, respectively. 
These parameters for the chosen examples are summarised in Table~\ref{tab:array_params}.

@@

Fast imaging involves processing data on multiple time scales. The channelisation of a voltage time stream of length $\delta t$ with a time resolution of $t_\textrm{s}$ will correspond to spectral channels of width $\delta\nu=1/\delta t$ and a bandwidth of $B=1/t_\textrm{s}$, respectively. The number of spectral channels is $N_\nu=B/\delta\nu=\delta t/t_\textrm{s}$. Images are expected to be produced at a cadence of $t_\textrm{acc}$, which is typically larger than $\delta t$ and smaller than the time scale on which calibration needs to be updated. In this paper, the nominal values chosen for $\delta t$ is 25~$\mu$s, and we explore the possibilities for $t_\textrm{acc}$ ranging from 100~$\mu$s to 10~s.

@@

There are other computations involved that may be relevant for different applications such as deconvolution, dedispersion, etc. that may determine the exact architecture.
